---
title: "Data/results not shown in dental research papers - analysis"
author: Eero Raittio, Ahmad Sofi-Mahmudi
date: "2022-02-10"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/ahmad/Google Drive/Meta-Research in Dentistry/Data Availability Project/data/')
```


# Loading required packages

```{r}
library(tidyverse)
```

# Loading the datasets
First, set working directory to the folder that has all the datasets in it using setwd() function and then run the following codes:

```{r}
db_all <- read.csv("db_all.csv")

db_dns <- read.csv("db_dns.csv")

hits_med_all <- read.csv("hits_med_all.csv")

hits_med_dns <- read.csv("hits_med_dns.csv")

```

# Results
## Overall perspective
Total number of papers in dental journals in all time:
```{r, echo=FALSE}
data.frame(number_of_papers = nrow(db_all),
           number_of_journals = nrow(table(db_all$journalTitle)))
```

Number of papers with "data/results not shown":
```{r, echo=FALSE}
data.frame(number_of_papers = nrow(db_dns),
           nummber_of_journals = nrow(table(db_dns$journalTitle)))
```


## Figures

First, adding the real publication year to the datasets. The real publication year is the year the paper was first appear online stored in firstPublicationDate column.
```{r}
library(lubridate)

db_all <- db_all %>%
  mutate(year_firstpub = year(
    as.POSIXlt(firstPublicationDate, 
               format = "%Y-%m-%d")
  ))

db_dns <- db_dns %>%
  mutate(year_firstpub = year(
    as.POSIXlt(firstPublicationDate, 
               format = "%Y-%m-%d")
  ))
```


Figure 1A (dental plot)
```{r}

trend_dent <- merge(count(db_dns, year_firstpub), count(db_all, year_firstpub), by = "year_firstpub")

colnames(trend_dent) <- c("year_firstpub", "data_not_shown", "all_fulltexts")

trend_dent <- mutate(trend_dent, pubYear_modified = 
                       ifelse(year_firstpub < 1990, "< 1990",
                     ifelse((year_firstpub >= 1990) & (year_firstpub < 1995), "1990 - 1994",
                     ifelse((year_firstpub >= 1995) & (year_firstpub < 2000), "1995 - 1999",
                     year_firstpub))))


trend_dent <- trend_dent %>% 
  group_by(pubYear_modified) %>% 
  summarise(data_not_shown = sum(data_not_shown),
            all_fulltexts = sum(all_fulltexts)) %>%
  mutate(percentage = round(data_not_shown/all_fulltexts*100, 1))


plot_dent <- trend_dent %>% 
  ggplot() +
  geom_col(aes(x = pubYear_modified, 
               y = data_not_shown),
           size = 1,
           fill = "cornflowerblue") +
  geom_line(aes(x = pubYear_modified, 
                y = 10*percentage), 
            size = 1.2,
            color="darkorange1", 
            linetype=2,group = 1) +
  geom_text(aes(label = data_not_shown, 
                x = pubYear_modified, 
                y = data_not_shown+3),
            color = "black", 
            size = 3.5) +
  geom_text(aes(label = percentage, 
                x = pubYear_modified, 
                y = 10*percentage), 
            color = "black", 
            size=3) +
  scale_y_continuous(sec.axis = sec_axis(~./10)) +
  theme_bw() + 
  theme(axis.title.x=element_blank(), axis.text.x = element_blank()) + 
  theme(axis.title.y=element_blank()) + 
  labs(title = "A", x = "Year")


plot_dent
```

Figure 1B (medical plot):
```{r}

trend_med <- merge(hits_med_dns, hits_med_all, by = c("year"))
trend_med <- trend_med[, -c(2, 3, 5, 6)]
colnames(trend_med) <- c("pubYear", "data_not_shown", "all_fulltexts")

trend_med <- mutate(trend_med, pubYear_modified = 
                       ifelse(pubYear < 1990, "< 1990",
                     ifelse((pubYear >= 1990) & (pubYear < 1995), "1990 - 1994",
                     ifelse((pubYear >= 1995) & (pubYear < 2000), "1995 - 1999",
                     pubYear))))


trend_med <- trend_med %>% 
  group_by(pubYear_modified) %>% 
  summarise(data_not_shown = sum(data_not_shown),
            all_fulltexts = sum(all_fulltexts)) %>%
  mutate(percentage = round(data_not_shown/all_fulltexts*100, 1))


plot_med <- trend_med %>% 
  ggplot() +
  geom_col(aes(x = pubYear_modified, 
               y = data_not_shown),
           size = 1,
           fill = "cornflowerblue") +
  geom_line(aes(x = pubYear_modified, 
                y = 2000*percentage), 
            size = 1.2,
            color="darkorange1", 
            linetype=2,group = 1) +
  geom_text(aes(label = percentage, 
                x = pubYear_modified, 
                y = 2000*percentage), 
            color = "black", 
            size=3) +
  scale_y_continuous(sec.axis = sec_axis(~./2000)) +
  guides(x =  guide_axis(angle = 45), y = guide_axis(angle = 74)) +
  theme_bw() + 
  theme(axis.title.y=element_blank()) +
  labs(title = "B", x = "Year")


plot_med

```

Combining plots:
```{r}
library(ggpubr)

figure <- ggarrange(plot_dent, plot_med + font("x.text", size = 10),
                    ncol = 1, nrow = 2)

anfig <- annotate_figure(figure,
                left = text_grob("The number of articles with data not shown", color = "black", rot = 90),
                right = "Percentage",
)


# tiff("Figure.tif", units = "cm", height = 31, width = 20, res = 300)
anfig
# dev.off()
```


## 2021 perspective
Total hits and journals for 2021:
```{r}
data.frame(total_hits = nrow(
  filter(db_all, 
         grepl("2021", firstPublicationDate))),
           total_journals = length(
  unique(
  filter(db_all, 
         grepl("2021", 
               firstPublicationDate))
  $journalTitle))
  
           )
```

For analysing publishers, we'll load publishers dataset:
```{r}

publishers <- read.csv("publishers.csv")
```


Join publishers to dataframes:
```{r}
db_all <- merge(db_all, publishers, by.x = "journalTitle", by.y = "ï..Journal", all.x = T)

db_dns <- merge(db_dns, publishers, by.x = "journalTitle", by.y = "ï..Journal", all.x = T)

```

Number of papers, journals and publishers that published papers with "data/results not shown" in 2021:
```{r}
data.frame(
  total_papers = nrow(filter(db_dns, grepl("2021", firstPublicationDate))),
  percentage = round(nrow(filter(db_dns, grepl("2021", firstPublicationDate)))/nrow(filter(db_all, grepl("2021", firstPublicationDate)))*100, 1),
  total_journals = length(unique(filter(db_dns, grepl("2021", firstPublicationDate))$journalTitle)),
  total_publishers = length(unique(filter(db_dns, grepl("2021", firstPublicationDate))$Publons))
)
```

Fisher's exact test for journal-related differences in publishing data not shown papers in 2021:
```{r}
library(data.table)

round(fisher.test(merge(
  table(db_all$journalTitle[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$journalTitle[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = TRUE) %>% mutate(Freq.y = replace(Freq.y, is.na(Freq.y), 0)) %>% select(Freq.x, Freq.y),
  simulate.p.value=TRUE,
  hybrid=T,
  B=100000
  )$p.value, 3)

```

Fisher's exact test for publisher-related differences in publishing data not shown papers in 2021 (based on Publons):
```{r}
round(fisher.test(merge(
  table(db_all$Publons[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$Publons[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = TRUE) %>% mutate(Freq.y = replace(Freq.y, is.na(Freq.y), 0)) %>% select(Freq.x, Freq.y),
  simulate.p.value=TRUE,
  hybrid=T,
  B=100000
  )$p.value, 3)
```

Fisher's exact test for publisher-related differences in publishing data not shown papers in 2021 (based on NLM Catalog):
```{r}
round(fisher.test(merge(
  table(db_all$NLMCatalog[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$NLMCatalog[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = TRUE) %>% mutate(Freq.y = replace(Freq.y, is.na(Freq.y), 0)) %>% select(Freq.x, Freq.y),
  simulate.p.value=TRUE,
  hybrid=T,
  B=100000
  )$p.value, 3)
```

Fisher's exact test for publisher-related differences in publishing data not shown papers in 2021 (based on SCImago):
```{r}
round(fisher.test(merge(
  table(db_all$SCImago[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$SCImago[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = TRUE) %>% mutate(Freq.y = replace(Freq.y, is.na(Freq.y), 0)) %>% select(Freq.x, Freq.y),
  simulate.p.value=TRUE,
  hybrid=T,
  B=100000
  )$p.value, 3)
```

## Tables
Table 2. The number of available open access articles from Europe PubMed Central and the number of articles with “data/results not shown” phrases in each journal:
```{r}

merge(
  table(db_all$journalTitle[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$journalTitle[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = T) %>% 
  mutate(proportion = round((Freq.y/Freq.x)*100, 1)) %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("journal", "total_studies", "studies_with_data_not_shown", "proportion"))
  
```

For "Others" row of Table 2:
```{r}
merge(
  table(db_all$journalTitle[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$journalTitle[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = T) %>%
  filter(is.na(Freq.y)) %>%
  summarise(total_journals = n(),
            total_papers = sum(Freq.x))
```

Table 3. The number of available open access articles from Europe PubMed Central and the number of articles with “data/results not shown” phrases in journals of each publisher:
```{r}

merge(
  table(db_all$Publons[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$Publons[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1") %>% 
  mutate(proportion = round((Freq.y/Freq.x)*100, 1)) %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("publisher", "total_studies", "studies_with_data_not_shown", "proportion"))
  
```

For "Others" row of Table 3:
```{r}
merge(
  table(db_all$Publons[db_all$firstPublicationDate %like% "2021"]) %>% as.data.frame(), 
  table(db_dns$Publons[db_dns$firstPublicationDate %like% "2021"]) %>% as.data.frame(),
  by = "Var1", all = T) %>%
  filter(is.na(Freq.y)) %>%
  summarise(total_publishers = n(),
            total_papers = sum(Freq.x))
```

## Analysing "data/results not shown" types in 2021
First loading dataset that contains categories for data not shown phrases in 2021 papers:
```{r}

dns_2021 <- read.csv("dns_2021.csv")

```

Frequency and proportion of each category:
```{r}
merge(
  table(dns_2021[, 11]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 11]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("result_type", "frequency", "percentage"))
```

Frequency and proportion of times they used the phrase in their paper:
```{r}
merge(
  table(dns_2021[, 13]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 13]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("times", "frequency", "percentage"))
```

Frequency and proportion of data and code sharing statement:
```{r}
merge(
  table(dns_2021[, 7]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 7]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("statement", "frequency", "percentage"))
```

Frequency and proportion of papers that shared data:
```{r}
merge(
  table(dns_2021[, 8]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 8]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("shared_data", "frequency", "percentage"))
```

Frequency and proportion of papers that shared code:
```{r}
merge(
  table(dns_2021[, 9]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 9]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("shared_code", "frequency", "percentage"))
```

Frequency and proportion of papers with supplementary materials:
```{r}
merge(
  table(dns_2021[, 10]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 10]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("supplementary", "frequency", "percentage"))
```

Frequency and proportion of papers with supplementary materials:
```{r}
merge(
  table(dns_2021[, 10]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 10]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("supplementary", "frequency", "percentage"))
```

Frequency and proportion of journals with unlimited supplementary material upload:
```{r}
merge(
  table(dns_2021[, 14]) %>% as.data.frame(),
  round(prop.table(table(dns_2021[, 14]))*100, 1) %>% as.data.frame(),
  by = "Var1") %>% 
  arrange(desc(Freq.y)) %>%
  setNames(c("status", "frequency", "percentage"))

```

Journals with unclear supplementary material upload status:
```{r}
dns_2021 %>%
  filter(across(14) == "Unclear") %>%
  select(6)
```

